%!TEX root = bloom_logical.tex

\section{Garbage Collection for Asynchronous Rules}
\label{sec:async_gc}
Consider an asynchronous merge rule on a sender node $v$
\begin{align*}
\overline{\tt{B}} \quad\tt{<}\sim\quad \tt{f}\tt{(}\tt{A}_1,\dots,\tt{A}_n\tt{)}
\end{align*}
and the corresponding merge rule on the receiver node $v'$
\begin{align*}
\tt{B} \quad\tt{<=}\quad \overline{\tt{B}}.
\end{align*}
We would like to garbage collect from $\tt{A}_1, \dots, \tt{A}_n$ and also to reduce the amount of communication needed.
Unfortunately, the asynchronous merge differs from a merge $\tt{B} ~\tt{<=}~ \tt{f}\tt{(}\tt{A}_1,\dots,\tt{A}_n\tt{)}$ or a deferred merge $\tt{B} ~\tt{<+}~ \tt{f}\tt{(}\tt{A}_1,\dots,\tt{A}_n\tt{)}$ in an important way:
the sender node has no knowledge of the contents of $\tt{B}$.

To fix this, \cite{conway2014edelweiss} introduced the following additional rules.
On the receiver side, we acknowledge all received messages with
\begin{align*}
\overline{\underline{\tt{B}}} \quad\tt{<}\sim\quad \overline{\tt{B}},
\end{align*}
which is stored on the sender side
\begin{align*}
\underline{\tt{B}} \quad\tt{<=}\quad \overline{\underline{\tt{B}}}.
\end{align*}

Thus, $\underline{\tt{B}}$ contains a subset of $\tt{B}$'s contents.
\begin{lemma}
\label{lem:async_merge_contains}
For any $(v,t) <_\tau (v',t')$, $I^{\Pi_\tau}_{v,t}(\underline{\tt{B}}@(v,t)) \subseteq I^{\Pi_\tau}_{v',t'}(\tt{B}@(v',t'))$.
\end{lemma}
\begin{proof}(Sketch)
For any $x \in I^{\Pi_\tau}_{v,t}(\underline{\tt{B}}@(v,t)) = \bigcup_{s=0}^t I^{\Pi_\tau}_{v,s}(\overline{\underline{\tt{B}}}@(v,s))$, it must be the case that $\exists s \leq t$: $x \in I^{\Pi_\tau}_{v,s}(\overline{\underline{\tt{B}}}@(v,s))$, i.e., $x$ was received via $\overline{\underline{\tt{B}}}$ at some earlier time $(v,s)$.
That in turn implies
$\exists t''$ such that $(v',t'') <_\tau (v,s)$ and $x \in I^{\Pi_\tau}_{v',t''}(\tt{B}@(v',t''))$, i.e., $x$ was sent from $v'$ at time $(v',t'')$.
Now $(v',t'') <_\tau (v,s) <_\tau (v,t) <_\tau (v',t')$ and $\tt{B}$ is persistent implies that $x \in I^{\Pi_\tau}_{v',t'}(\tt{B}@(v',t'))$.
\end{proof}

\subsection{Logical Garbage Collection}
Armed with (partial) knowledge of $\tt{B}$, we can now introduce logical garbage collection for the asynchronous merge rule.
On the sender node $v$ in $\mathfrak{P}_{GC}$, we have rules
\begin{align}
(\overline{\tt{B}},\emptyset) &\quad\tt{<}\sim\quad (\tt{f}\tt{(}\tt{A}_{1,\exists} \cup \tt{A}_{1,\top}, \dots, \tt{A}_{n,\exists} \cup \tt{A}_{n,\top}\tt{)}, \emptyset)\label{eq:async_logical_send}\\
(\underline{\tt{B}}_\exists, \underline{\tt{B}}_\top) &\quad\tt{<=}\quad (\overline{\underline{\tt{B}}}, \emptyset),\nonumber
\end{align}
and on the receiver node $v'$,
\begin{align*}
(\tt{B}_\exists, \tt{B}_\top) &\quad\tt{<=}\quad (\overline{\tt{B}}, \emptyset)\\
(\overline{\underline{\tt{B}}}, \emptyset) &\quad\tt{<}\sim\quad (\overline{\tt{B}}, \emptyset).
\end{align*}
Note that we can replace \eqref{eq:async_logical_send} with $(\overline{\tt{B}},\emptyset) ~\tt{<}\sim~ (\tt{f}\tt{(}\tt{A}_{1,\exists} \cup \tt{A}_{1,\top}, \dots, \tt{A}_{n,\exists} \cup \tt{A}_{n,\top}\tt{)}  - (\underline{\tt{B}}_\exists \cup \underline{\tt{B}}_\top), \emptyset)$ without affecting the overall program logic.
We term this trick ARM, avoidance of redundant messages.

In addition, we add the garbage collection rule
\begin{align}
\langle \tt{A}_{1,TS}, \dots, \tt{A}_{n,TS} \rangle
\quad\tt{<+}\quad
\tt{GC}(\tt{A}_{1,TS}, \dots, \tt{A}_{n,TS}; \underline{\tt{B}}_{TS})
\end{align}
which depends on the $v$-accessible $\underline{\tt{B}}_{TS}$ instead of $\tt{B}_{TS}$.

Analogously to Merge GC Invariant \ref{inv:merge_gc}, we will require that the Asynchronous Merge GC Invariant \ref{inv:async_merge_gc} is maintained by all rules, including the above GC rule.

\begin{invariant}[Asynchronous Merge GC Invariant]\label{inv:async_merge_gc}
Consider any asynchronous merge rule 
\[R: \tt{B} ~\tt{<}\sim~ \tt{f(}\tt{A}_1,\dots,\tt{A}_n, \tt{U}_1, \dots, \tt{U}_m\tt{)}\]
where
$\tt{A}_1,\dots,\tt{A}_n$ are tables without keys,
$\tt{U}_1,\dots,\tt{U}_m$ are tables with keys,
and RHS is not a channel.
We require that, 
$\forall \langle \widehat{\tt{A}}_1, \dots, \widehat{\tt{A}}_n\rangle \geq \langle \tt{A}_{1,\exists},\dots,\tt{A}_{n,\exists} \rangle,
\quad
\forall \langle \widehat{\tt{U}}_1 , \dots, \widehat{\tt{U}}_m \rangle \geq \langle \tt{U}_{1,\exists}, \dots, \tt{U}_{m,\exists} \rangle$
such that $\emptyset = \widehat{\tt{U}}_1 \cap \tt{U}_{1,\top} = \dots = \widehat{\tt{U}}_1 \cap \tt{U}_{m,\top}$:
\begin{align}
\tt{f(}\widehat{\tt{A}}_1,\dots,\widehat{\tt{A}}_n, \widehat{\tt{U}}_{1,\exists},\dots,\widehat{\tt{U}}_{n,\exists}\tt{)}
\cup \underline{\tt{B}}_\exists\cup \underline{\tt{B}}_\top
~=~
\tt{f(}
\widehat{\tt{A}}_1 \cup \tt{A}_{1,\top},\dots,\widehat{\tt{A}}_n \cup \tt{A}_{n,\top}, \widehat{\tt{U}}_1 \cup \tt{U}_{1,\top},\dots,\widehat{\tt{U}}_n \cup \tt{U}_{n,\top}\tt{)} 
\cup \underline{\tt{B}}_\exists \cup \underline{\tt{B}}_\top.
\label{eq:inv_async_merge_gc}
\end{align}
\end{invariant}
If \eqref{eq:inv_async_merge_gc} holds for $\underline{\tt{B}}_{TS}@(v,t)$, then by Lemma \ref{lem:async_merge_contains}, it also holds for $\tt{B}_{TS}@(v',t')$ if $(v,t) <_\tau (v',t')$.

Correspondingly, we say a GC rule is \emph{safe for} an asynchronous rule $R$ if it maintains the Asynchronous Merge Invariant \ref{inv:async_merge_gc} with respect to $R$.

\begin{property}[Asynchronous Merge GC Safety]
A garbage collection rule $\tt{GC}$ is safe for an asynchronous merge rule $R: \tt{B} ~\tt{<}\sim~ \tt{f(}\tt{A}_1,\dots,\tt{A}_n, \tt{U}_1, \dots, \tt{U}_m\tt{)}$ if it respects output, and maintains the Asynchronous Merge GC Invariant \ref{inv:async_merge_gc} for $R$.
That is, if $\langle \tt{A}_{1,TS}, \dots, \tt{A}_{n,TS}, \tt{U}_{1,TS}, \dots, \tt{U}_{m,TS} \rangle$ and $\underline{\tt{B}}_{TS}$ satisfies Asynchronous Merge GC Invariant \ref{inv:async_merge_gc}, then
$\langle \tt{A}_{1,TS}^*, \dots, \tt{A}_{n,TS}^*,$ $\tt{U}_{1,TS}^*, \dots, \tt{U}_{m,TS}^* \rangle$ $=$ $\tt{GC}(\tt{A}_{1,TS}, \dots, \tt{A}_{n,TS},$ $\tt{U}_{1,TS}, \dots, \tt{U}_{m,TS}; \underline{\tt{B}}_{TS})$ and $\underline{\tt{B}}_{TS}$ satisfies Asynchronous Merge GC Invariant \ref{inv:async_merge_gc}.
\end{property}

Unlike with the Merge GC Safety Property \ref{property:gc_safety}, we require that $\tt{GC}$ maintains Invariant \ref{inv:async_merge_gc} without the condition that 
$\tt{B} \supseteq \tt{f}\tt{(}\tt{A}_{1,\exists} \cup \tt{A}_{1,\top},\dots,\tt{A}_{n,\exists} \cup \tt{A}_{n,\top}\tt{)}$,
since there is no guarantee that this condition is true at the time that the GC rule is executed, nor that it will ever be true if an unreliable messaging protocol is used.
As a simple example, consider the asynchronous copy rule $\tt{B} \tt{<}\sim \tt{A}$.
If we apply $\tt{GCAll}$ for the linear copy function, we could tombstone all tuples in $\tt{A}$ and never send them again.
However, the message could be dropped and thus never be seen by $\tt{B}$.


We will require that the garbage collection rule $\tt{GC}$ that is added to $\mathfrak{P}_{GC}$ for asynchronous merge rule $R$ be safe for $R$, monotone, and conservative.

It is not difficult to see that $\mathfrak{P}_{GC}$ with garbage collection for asynchronous merges maintains both Merge GC Invariant \ref{inv:merge_gc} and Asynchronous Merge GC Invariant \ref{inv:async_merge_gc}.
We omit the proof as it largely follows the structure of the proof of Lemma \ref{lem:merge_gc_inv}.

\begin{lemma}
\label{lem:async_merge_gc_inv}
The rewritten GC program $\mathfrak{P}_{GC}$ maintains the Merge GC Invariant \ref{inv:merge_gc} and the Asynchronous Merge GC Invariant \ref{inv:async_merge_gc}.
\end{lemma}

Furthermore, Theorem \ref{thm:equivalence_of_sets}, Corollary \ref{cor:equivalence_of_output}, Theorem \ref{thm:coord-free}, and Corollary \ref{cor:confluence_of_output} hold with little modification of the proofs.


\subsection{Instantiated Garbage Collection}
On the sender node $v$ in $\mathfrak{P}_{iGC}$, we have rules
\begin{align*}
\overline{\tt{B}}_I &\quad\tt{<}\sim\quad \tt{f}\tt{(}\tt{A}_{1,I}, \dots, \tt{A}_{n,I}\tt{)} - \underline{\tt{B}}_I\\
\underline{\tt{B}}_I &\quad\tt{<=}\quad \overline{\underline{\tt{B}}}_I\\
\langle \tt{A}_{1,E}, \dots, \tt{A}_{n,E} \rangle
&\quad\tt{<-}\quad
\tt{GCI}(\tt{A}_{1,E}, \dots, \tt{A}_{n,E}; \underline{\tt{B}}_{E})
\end{align*}
and on the receiver node $v'$,
\begin{align*}
\tt{B}_I &\quad\tt{<=}\quad \overline{\tt{B}}_I & \text{(or $(\tt{B}_I, \tt{B}_!) \quad\tt{<=}\quad (\overline{\tt{B}}_I, \pi_!(\overline{\tt{B}}_I$))}\\
\overline{\underline{\tt{B}}}_I &\quad\tt{<}\sim\quad \overline{\tt{B}}_I.
\end{align*}

As before, we require that $\tt{GCI}$ is conservative and consistent with some $\tt{GC}$ which is itself safe for $\tt{R}$, monotone, and conservative.

The proof of Theorem \ref{thm:instantiated_correctness} is largely unchanged, except with some careful handling of the differences between $\overline{\tt{B}}_I$ and $\underline{\tt{B}}$ resulting from the instantiated garbage collection and ARM.
The implications of Corollaries \ref{cor:instantiated_equivalence_of_output}, \ref{cor:instantiated_confluence_of_output}, \ref{cor:instantiated_confluence_of_keys} follow from Theorem \ref{thm:instantiated_correctness} without change.



\subsection{Garbage Collecting $\underline{\tt{B}}$}
We assume WLOG in this subsection that the RHS of an asynchronous rule is a table $\tt{A}$ with primary keys.
If not, we can replace $\overline{\tt{B}} ~\tt{<}\sim~ \tt{f}\tt{(}\tt{A}_1, \dots, \tt{A}_n\tt{)}$ with rules $\overline{\tt{B}} ~\tt{<}\sim~ \tt{A}$ and $\tt{A} ~\tt{<=}~ \tt{f}\tt{(}\tt{A}_1, \dots, \tt{A}_n\tt{)}$, where we attach unique message IDs to tuples merged into $\tt{A}$.

Then, we can apply DR- to the RHS $\tt{A}_I - \underline{\tt{B}}_I$ to garbage collect from both $\tt{A}_E$ and $\underline{\tt{B}}_E$, by replacing the garbage collection rule with
\[
\langle \tt{A}_E, \underline{\tt{B}}_E \rangle \quad\tt{<-}\quad \tt{iGCDR-}(\tt{A}_E, \underline{\tt{B}}_E; \underline{\tt{B}}_E).
\]
